---
title: "Jeopardy Analysis"
author: "Elizabeth Wickes"
date: "October 27, 2014"
output: html_document
---

```{r echo=FALSE}
options("scipen"=999)

printMoney <- function(x){
  format(round(x,digits=2), digits=10, nsmall=2, decimal.mark=".", big.mark=",")
}


```


```{r echo =FALSE}

usplayerswithsoc <- read.csv("~/GoogleDrive/_SODA/JeopardyDOLData/usplayerswithsocNODUPEdivreg.csv")

# replace empty values with NA and assign to pretty names

usplayerswithsoc$id [usplayerswithsoc$ID == ''] <- NA
id <- usplayerswithsoc$ID

usplayerswithsoc$name [usplayerswithsoc$name == ''] <- NA
name <- usplayerswithsoc$name

usplayerswithsoc$year [usplayerswithsoc$year== ''] <- NA
year <- usplayerswithsoc$year

usplayerswithsoc$occ [usplayerswithsoc$occ == ''] <- NA
occ <- usplayerswithsoc$occ

usplayerswithsoc$loc [usplayerswithsoc$loc == ''] <- NA
loc <- usplayerswithsoc$loc

usplayerswithsoc$numgames [usplayerswithsoc$numgames == ''] <- NA
numgames <- usplayerswithsoc$numgames

usplayerswithsoc$totalwinnings [usplayerswithsoc$totalwinnings == ''] <- NA
winnings <- usplayerswithsoc$totalwinnings

usplayerswithsoc$state [usplayerswithsoc$state == ''] <- NA
state <- usplayerswithsoc$state

usplayerswithsoc$studentflag [usplayerswithsoc$studentflag == ''] <- NA
studentflag <- usplayerswithsoc$studentflag

usplayerswithsoc$SOC [usplayerswithsoc$SOC == '' | usplayerswithsoc$SOC == '9999'] <- NA
soc <- usplayerswithsoc$SOC

usplayerswithsoc$main [usplayerswithsoc$main == '' | usplayerswithsoc$main == '9999'] <- NA
main <- usplayerswithsoc$main

usplayerswithsoc$sub [usplayerswithsoc$sub == ''] <- NA
usplayerswithsoc$sub [usplayerswithsoc$main == '' | usplayerswithsoc$sub == '9999'] <- NA
sub <- usplayerswithsoc$sub

usplayerswithsoc$region [usplayerswithsoc$region == ''] <- NA
region <- usplayerswithsoc$region

usplayerswithsoc$division [usplayerswithsoc$division == ''] <- NA
division <- usplayerswithsoc$division


```

``` {r echo =FALSE}


# make cleanplayers table

cleanplayers <- data.frame(id, name, year, occ, loc, numgames, winnings, state, soc, main, sub, region, division)

# Convert quantitative variables to numeric
cleanplayers$year <- as.numeric(cleanplayers$year)
cleanplayers$winnings <- as.numeric(cleanplayers$winnings)
cleanplayers$numgames <- as.numeric(cleanplayers$numgames)

# replace possible bad values of years with NA
# this shouldn't be the case with the most current code,
# but possible that it may happen in the future.
cleanplayers$year [cleanplayers$year < 1950] <- NA
cleanplayers$year [cleanplayers$year > 2015] <- NA

#Create anoter version of the player data frame only containing players without NA values, and reassign variables.
#This reduced the total number of players for analysis to about 5,000.



cleanplayersnoNA <- cleanplayers

# I'm leaving the following code in, but it should be considered to be deprecated.    
# removing the line from comment below will remove all players with NA values from the dataset.
# NAs are being handled on a more elegant basis, but it may be of use later.

#cleanplayersnoNA <- na.omit(cleanplayers)

id <- cleanplayersnoNA$id
name <- cleanplayersnoNA$name
year <- cleanplayersnoNA$year
occ <- cleanplayersnoNA$occ
loc <- cleanplayersnoNA$loc
numgames <- cleanplayersnoNA$numgames
winnings <- cleanplayersnoNA$winnings
state <- cleanplayersnoNA$state
studentflag <- cleanplayersnoNA$studentflag
soc <- cleanplayersnoNA$soc
main <- cleanplayersnoNA$main
sub <- cleanplayersnoNA$sub
divsion <- cleanplayersnoNA$division
region <- cleanplayersnoNA$region

# create the 2003 and after set

after2003 <- subset(cleanplayersnoNA, year >= 2003)

after2003noken <- subset(after2003, id != '661')
noken <- subset(cleanplayersnoNA, id != '661')

kenswinnings <- after2003$winnings[after2003$id == '661']
after2003sd <- sd(after2003$winnings)
after2003nokensd <- sd(after2003noken$winnings)

cleanplayersnoNA$after2003 <- NA
cleanplayersnoNA$after2003 [cleanplayersnoNA$year >= 2003] <- "2003 and after"
cleanplayersnoNA$after2003 [cleanplayersnoNA$year < 2003] <- "Before 2003"


justone <- subset(cleanplayersnoNA, cleanplayersnoNA$numgames == 1)
morethanone <- subset(cleanplayersnoNA, cleanplayersnoNA$numgames > 1)

justoneafter2003 <- subset(after2003, after2003$numgames == 1)
morethanoneafter2003 <- subset(after2003, after2003$numgames > 1)
morethanoneafter2003noken <- subset(after2003noken, after2003noken$numgames > 1)

```

# Data description

### Collection and preparation 

There were a total of `r length(usplayerswithsoc$ID)` players were imported, representing the subset of players who were detected to be from the United States and all duplicates removed.  Duplicates were defined as entities with matching name and winnings values.

Data were pulled from player statistics HTML pages from [J-Archive!](http://j-archive.com/).  Example page:  
[http://j-archive.com/showplayerstats.php?player_id=8000](http://j-archive.com/showplayerstats.php?player_id=8000).

Players were scraped in and associated data gathered.  Occupational statements were aligned to the [2010 SOC code system](http://www.bls.gov/soc/major_groups.htm).  28% of the players are failing to match with a controlled SOC code.  Some additional SOC codes and groups were added to increase this matching.  The curated version of the SOC code system is available [here](https://github.com/elliewix/JeopardyDOLData/blob/master/scripts-soccodes.csv).

All code to produce the associated CSV files is located in an IPython Notebook available for viewing and download [here](http://nbviewer.ipython.org/github/elliewix/JeopardyDOLData/blob/master/PreparePlayerFiles.ipynb).  All original player and analysis files are available within [this github repository](http://nbviewer.ipython.org/github/elliewix/JeopardyDOLData/tree/master/).

The data have been split to only look at players participating in 2003 and after due a rules change that removed the maximum number of games each participant may compete in.  The frequencies of players participating in numbers of games are charted below.

### Exploring the impact of Ken Jennings

The majority of plots and other descriptives involving winnings will utilize a dataset that excludes Ken Jennings.  Ken is the highest winning player ever on Jeopardy!, with a total of $`r printMoney(kenswinnings)`.  The standard deviation for the set of players after 2003 with Ken is $`r printMoney(after2003sd)` and is $`r printMoney(after2003nokensd)` when Ken is removed.  His winnings have a z-score of `r (kenswinnings - mean(after2003$winnings) ) / sd(after2003$winnings)` when compared to the data set of including his value, and `r (kenswinnings - mean(after2003noken$winnings) ) / sd(after2003noken$winnings)` when excluding him.  This z-score goes down to `r (kenswinnings - mean(morethanoneafter2003$winnings)) / sd(morethanoneafter2003$winnings)` when looking just within the subset of players who have participated in more than one game.

| Data | Mean | SD |
| ---- | ---- | -- | 
| 2003 and after, all players | $`r printMoney(mean(after2003$winnings))` | $`r printMoney(sd(after2003$winnings))` |
| 2003 and after, excluding Ken | $`r printMoney(mean(after2003noken$winnings))` |  $`r printMoney(sd(after2003noken$winnings))` | 

The distribution of winnings is highly skewed to the right and lacks a normal distribution.  However, the distribution of winnings for players participating in just one game (indicating that they played without winning) is less skewed, but still not normal.  The data below include Ken Jennings.

```{r echo = F}

hist(after2003noken$winnings,breaks=50, main='Winnings histogram for all players after 2003\n(excluding Ken Jennings)',xlab='Winnings')

hist(justoneafter2003$winnings, breaks=50, main = 'Winnings histogram for players after 2003\ncompeting in only one game',xlab="Winnings")

hist(morethanoneafter2003$winnings, breaks=50, main = 'Winnings histogram for players\ncompeting in more than one game\n(including Ken Jennings)',xlab="Winnings")

hist(morethanoneafter2003noken$winnings, breaks=50, main = 'Winnings histogram for players\ncompeting in more than one game\n(excluding Ken Jennings)',xlab="Winnings")

```

```{r echo= F}

maxfreq <- function(var){
  df <- data.frame(table(var))
  max <-df[df$Freq == max(df$Freq), ]
  return (max)
}

popcount <- length(id)

```


# Descriptive statistics 
###Categorial values

| Variable | Value | Count | NAs |
| -------- | ----- | ----- | --- |
| Occupation phrase | `r maxfreq(occ)[1]` | `r maxfreq(occ)[2]` | `r length(occ[is.na(occ)])`  |
| Location phrase | `r maxfreq(loc)[1]` | `r maxfreq(loc)[2]` |  `r length(loc[is.na(loc)])` |
| Controlled state value | `r maxfreq(state)[1]` | `r maxfreq(state)[2]` | `r length(state[is.na(state)])` |
| SOC code | `r maxfreq(soc)[1]` | `r maxfreq(soc)[2]` | `r length(soc[is.na(soc)])`  |
| SOC Major code | `r maxfreq(main)[1]` | `r maxfreq(main)[2]` | `r length(main[is.na(main)])`  |
| United States division | `r maxfreq(division)[1]` | `r maxfreq(division)[2]` |  `r length(division[is.na(division)])` |
| United States region | `r maxfreq(region)[1]` | `r maxfreq(region)[2]` | `r length(region[is.na(region)])`  |

###Quantitative values

| Variable | Minimum | 1st Quartile | Median | Mean | 3rd Quartile | Maximum | NAs |
| -------- | ------- | ------------ | ------ | ---- | ------------ | ------- | --- |
| Numer of games participated in  | `r min(numgames, na.rm=TRUE)` | `r quantile(numgames, na.rm=TRUE)[2]` | `r median(numgames, na.rm=TRUE)`  |  `r mean(numgames, na.rm=TRUE)` |  `r quantile(numgames, na.rm=TRUE)[4]` |  `r max(numgames, na.rm=TRUE)` | `r length(numgames[is.na(numgames)])`  |
| Total winnings | `r min(winnings, na.rm=TRUE)` | `r quantile(winnings, na.rm=TRUE)[2]` | `r median(winnings, na.rm=TRUE)`  |  `r mean(winnings)` |  `r quantile(winnings, na.rm=TRUE)[4]` |  `r max(winnings, na.rm=TRUE)` | `r length(winnings[is.na(winnings)])` (False positives not being detected) |

### Summary of the 2003 subset

#### Winnings
| Data | n | Mean | Minimum | Maximum | SD |
| ---- | - | ---- | ------- | ------- | -- |
| All players after 2003 | `r length(after2003$id)` | $`r printMoney(mean(after2003$winnings))` | $`r printMoney(min(after2003$winnings))` | $`r printMoney(max(after2003$winnings))` | $`r printMoney(sd(after2003$winnings))` |
| Players only participating in one game | `r length(justoneafter2003$id)` | $`r printMoney(mean(justoneafter2003$winnings))` | $`r printMoney(min(justoneafter2003$winnings))` | $`r printMoney(max(justoneafter2003$winnings))` | $`r printMoney(sd(justoneafter2003$winnings))` |
| Players participating in more than one game\n (including Ken Jennings) | `r length(morethanoneafter2003$id)` | $`r printMoney(mean(morethanoneafter2003$winnings))` | $`r printMoney(min(morethanoneafter2003$winnings))` | $`r printMoney(max(morethanoneafter2003$winnings))` | $`r printMoney(sd(morethanoneafter2003$winnings))` |
| Players participating in more than one game\n (excluding Ken Jennings) | `r length(morethanoneafter2003noken$id)` | $`r printMoney(mean(morethanoneafter2003noken$winnings))` |  $`r printMoney(min(morethanoneafter2003noken$winnings))` |  $`r printMoney(max(morethanoneafter2003noken$winnings))` |  $`r printMoney(sd(morethanoneafter2003noken$winnings))` |

The $0 minimum values for the players participating in more than one game are likely artifacts of missing game data within the database.  Manual inspection of the data shows that there are only four of these cases.

#### Number of games participated in

| Variable | Minimum | 1st Quartile | Median | Mean | 3rd Quartile | Maximum | NAs |
| -------- | ------- | ------------ | ------ | ---- | ------------ | ------- | --- |
| Numer of games participated in  | `r min(after2003$numgames, na.rm=TRUE)` | `r quantile(after2003$numgames, na.rm=TRUE)[2]` | `r median(after2003$numgames, na.rm=TRUE)`  |  `r mean(after2003$numgames, na.rm=TRUE)` |  `r quantile(after2003$numgames, na.rm=TRUE)[4]` |  `r max(after2003$numgames, na.rm=TRUE)` | `r length(numgames[is.na(numgames)])`  |
| Total winnings | `r min(after2003$winnings, na.rm=TRUE)` | `r quantile(after2003$winnings, na.rm=TRUE)[2]` | `r median(after2003$winnings, na.rm=TRUE)`  |  `r mean(after2003$winnings)` |  `r quantile(after2003$winnings, na.rm=TRUE)[4]` |  `r max(after2003$winnings, na.rm=TRUE)` | `r length(winnings[is.na(winnings)])` (False positives not being detected) |


#Plots

Unless mentioned otherwise, all plots are using the data from after 2003.

```{r echo = F}
#barplot(table(after2003$main), horiz=T, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Frequencies of players \nwithin SOC major codes", xlab = "Frequency", ylab = "Major code")

barplot(table(cleanplayers$main), horiz=T, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Frequencies of players \nwithin SOC major codes\n(all player data)", xlab = "Frequency", ylab = "Major code")

#barplot(table(after2003$region), horiz=T, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Frequencies of players and region", xlab = "Frequency", ylab="United States census region")

barplot(table(cleanplayers$region), horiz=T, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Frequencies of players and region\n(all players)", xlab = "Frequency", ylab="United States census region")

barplot(table(after2003$division), horiz=T, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Frequencies of players and\nUnited States Census Bureau region\n(2003 and after)", xlab = "Frequency")

barplot(table(cleanplayers$division), horiz=T, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Frequencies of players and\nUnited States Census Bureau region\n(all players", xlab = "Frequency")

barplot(table(after2003$year), horiz=T, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Number of games in the\ndataset for each year 2003 and after", xlab = "Games", ylab = "Year")

```



```{r echo=FALSE}

after2003counts <- table(cleanplayersnoNA$after2003, cleanplayersnoNA$numgames)
barplot(after2003counts, main="Number of games each player\nparticipated in grouped by game era", ylab="Frequency", xlab="Number of games", col=c("cadetblue4","antiquewhite"),legend = rownames(after2003counts), beside = T)

```

###Exploration of winnings and categorical data

The data below are all excluding Ken Jennings.

```{r echo = F}
boxplot(after2003noken$winnings ~ after2003noken$region, horizontal=T, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Winnings by region", xlab = "Winnings", ylab = "Region")

boxplot(after2003noken$winnings ~ after2003noken$division, horizontal=T, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Winnings by division", xlab = "Winnings", ylab = "Division")

boxplot(after2003noken$numgames ~ after2003noken$region, horizontal=T, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Number of games played by players in regions", xlab = "Games played", ylab = "Region")
```

```{r echo = F}

#winningmeans <-aggregate(after2003noken$winnings, by=list(after2003noken$region, after2003noken$division), FUN=mean, na.rm = T)

divwinnings <- after2003noken$winnings
divdivision <- after2003noken$division
divagg <- data.frame(divwinnings, divdivision)
divmeans <- aggregate(divwinnings, by = list(divdivision), FUN=mean, na.rm=T)
colnames(divmeans) <- c("division", "mean")
barplot(divmeans$mean, names.arg=divmeans$division, horiz = T, cex.names=0.5, cex.axis=.5, las=1, xlab = "Mean winnings", ylab = "United States divisions", main="Mean winnings by division")

mainwinnings <- after2003noken$winnings
mainmain <- after2003noken$main
mainagg <- data.frame(mainwinnings, mainmain)
mainmeans <- aggregate(mainwinnings, by = list(mainmain), FUN=mean, na.rm=T)
colnames(mainmeans) <- c("majorcode", "mean")
barplot(mainmeans$mean, names.arg=mainmeans$majorcode, horiz = T, cex.names=0.5, cex.axis=.5, las=1, xlab = "Mean winnings", ylab = "SOC Major codes", main="Mean winnings by SOC Major code")

df <- data.frame(table(after2003noken$main))
df <- subset(df, df$Freq > 0)
dffreq <- df$Freq
dfmeainmeans <- mainmeans
dfmeans <- data.frame(mainmeans, dffreq)
dfmeans <- subset(dfmeans, dfmeans$dffreq > 5)
morethanfivewinnings <- dfmeans$mean
morethanfivemains <- dfmeans$majorcode
morethanfiveagg <- data.frame(morethanfivewinnings, morethanfivemains)
morethanfivemainmeans <- aggregate(morethanfivewinnings, by = list(morethanfivemains), FUN=mean, na.rm=T)
colnames(morethanfivemainmeans) <- c("majorcode", "mean")
barplot(morethanfivemainmeans$mean, names.arg=morethanfivemainmeans$majorcode, horiz = T, cex.names=0.5, cex.axis=.5, las=1, xlab = "Mean winnings", ylab = "SOC Major codes", main="Mean winnings by SOC Major code\n with more than 5 instances")

```

Several of the SOC major codes only have 1-5 players, so a subset of tabulations was created for better comparision between the groups.

```{r echo = F}
 #winningagg <- aggregate(cleanperson, by = list( main = justwinnings$cleanplayersnoNA.main), mean)

#dotchart(winningagg$x, labels = winningagg$main, cex=.7)
#playersagg <- data.table(cleanplayersnoNA.main, cleanplayersnoNA.winnings)
#playersagg <- as.data.frame(playersagg[, mean(cleanplayersnoNA.winnings, na.rm = TRUE),by = cleanplayersnoNA.main])

#aggmain <- playersagg$cleanplayersnoNA.main
#aggwinavg <- playersagg$V1

#dotchart(aggwinavg, labels = aggmain, cex=.7)
```


### Plots showing the impact of Ken Jennings

```{r echo = F}

noken <- subset(cleanplayersnoNA, id != '661')

boxplot(cleanplayersnoNA$winnings ~ cleanplayersnoNA$main, data = cleanplayersnoNA, main = "Winnings by SOC major code\n(including Ken Jennings)", xlab = "SOC major code", ylab = "Winnings")

boxplot(noken$winnings ~ noken$main, data = noken, main = "Winnings by SOC major code\n(excluding Ken Jennings)", xlab = "SOC major code", ylab = "Winnings")

```


### Winnings over year of the show

```{r echo =FALSE}

boxplot(noken$winnings ~ noken$year, data = noken, main = "Winnings over time", xlab = "Year", ylab = "Winnings")
boxplot(after2003noken$winnings ~ after2003noken$year, data = noken, main = "Winnings over time", xlab = "Year", ylab = "Winnings")

```
  

```{r echo = F}

barplot(table(cleanplayersnoNA$main, cleanplayersnoNA$numgames), col=c("darkblue","red"), horiz=F, cex.names=0.5, cex.axis=.5, beside=T, las=1, main="Number of games won per player", xlab = "Games", ylab = "Player frequency")

boxplot(justone$winnings ~ justone$main, data = justone)

boxplot(morethanone$winnings ~ morethanone$state, data = morethanone)

nokenmroethanone <- subset(morethanone, id != '661')

boxplot(nokenmroethanone$winnings ~ nokenmroethanone$main, data = nokenmroethanone, format)

```

Frequency of players within major SOC codes and game winners or not, using 2003 and after data.

```{r echo = F}

cleanplayersnoNA$gameswoncat <- "Won 1+"
cleanplayersnoNA$gameswoncat [cleanplayersnoNA$numgames == 1] <- "Won none"

counts <- table(cleanplayersnoNA$gameswoncat, cleanplayersnoNA$main)

barplot(counts, main="Frequency of players within major SOC coes", ylab="Frequency", xlab="Major code", col=c("cadetblue4","antiquewhite"),legend = rownames(counts))


```

There are a total of `r length(cleanplayersnoNA$id)` players remaining within the dataset after omitting those with missing values.  

http://www.inside-r.org/packages/cran/epicalc/docs/aggregate.plot

play2013 <- subset(cleanplayersnoNA, year == 2013)

boxplot(play2013$winnings ~ play2013$main, data = play2013)

barplot(table(play2013$winnings, play2013$main))

